The importance of a complete statistical description of dependence between variables for ecology and related fields

### 2019-04-06 Paper.Rmd
answering Reid's comment and adding notes for DAN

---
title: "The importance of a complete statistical description of dependence between variables for ecology and related fields"
author: "Shyamolina Ghosh, Lawrence W. Sheppard, Mark T. Holder, Terrance E. Loecke, P. Chris Reid, Daniel C. Reuman"
date: ""
geometry: "left=1.5cm,right=1.5cm,top=2.5cm,bottom=2.8cm"

@@

Reid: Marine Institute, Plymouth University, Drake Circus, Plymouth, United Kingdom; Mrine Biological Association of the UK, The Laboratory, Citadel Hill, Plymouth, United Kingdom

@@

Very generally, all branches of ecology and related areas study relationships among environmental and biological 
variables. However, commonly used correlation and regression 
approaches to studying 
such relationships are limited - common statistical
approaches provide only a very partial description of a relationship between variables. 
For instance, datasets showing markedly different 
relationships may nevertheless have the same correlation coefficient value 
(Fig. \ref{fig_cop_pedag1}A-B). The variables of Fig. \ref{fig_cop_pedag1}A 
(respectively, Fig. \ref{fig_cop_pedag1}B) are principally 
related in the left (respectively, right) portions of their distributions, an asymmetric 
pattern of dependence that may
have ecological significance, as discussed below, but is not captured by 
correlation. Panels C and D 
were obtained from A and B through monotonic transformation 
(Appendix \ref{SM-pedagog1_details} for details),
which alters the Pearson, but not the Spearman correlation. Although correlation
is only one of numerous metrics for 
studying relationships between variables, other common metrics also 
provide only partial information.

@@

Well-developed statistical tools do exist, and are 
applied widely in other fields, that provide a demonstrably complete 
description of the dependence 
between two or more variables [@joe2014_dependence; @nelsen2006_copula]. 
The goal of this paper is to illustrate the potential for 
broader application of these
tools in ecology and related fields. The tools are based on 
the statistical concept of the *copula*, which we here 
introduce. Copula approaches separate the information contained in a bivariate 
random variable into two distinct parts: the information contained in the 
marginal distributions (which says nothing about the dependence between the 
variables), and the remaining information (which is solely about
the dependence between the variables). From a data perspective, given 
a sample $(x_i,y_i)$, $i=1,2,\ldots,n$ (e.g., 
Fig. \ref{fig_cop_pedag1}A-B), the associated dependence
or copula structure is revealed by the bivariate plot of $u_i$ versus $v_i$, where
$u_i$ is the rank of $x_i$ in the set of the $x_j$ ($j=1,\ldots,n$), divided by
$n+1$; and $v_i$ is defined in the same way but using the $y_i$. See, 
for instance,
Fig. \ref{fig_cop_pedag1}E-F, which show the copula structure of 
A-B, respectively. Ranking serves to uniformize the 
marginal distributions of the underlying random variables, isolating the 
dependence structure only.
Mathematically, if $(X,Y)$ is a continuous bivariate random
variable and the cumulative distribution functions (cdfs) of $X$ and $Y$ are 
$F_X(x)=P[X\leq x]$ and $F_Y(y)=P[Y\leq y]$, then the dependence or copula structure
of $(X,Y)$ is revealed by the random variable $(F_X(X),F_Y(Y))$.
This random variable takes values in the unit square, 
$[0,1] \times [0,1]$, and has uniform marginal distributions. Probability density 
functions (pdfs) corresponding to the copulas associated 
with Fig. \ref{fig_cop_pedag1}E-F are in Fig.
\ref{fig_cop_pedag1}G-H, respectively. 
See @nelsen2006_copula and @joe2014_dependence
for general background. An theorem of A. Sklar 
states that copulas give a unique and complete 
representation of the dependence
between variables, in the sense that the joint distribution of $(X,Y)$ 
can be reconstructed from the 
copula and the marginal distributions [@sklar1959_theorem]. 
Copula theory is also well developed for
multivariate distributions.

@@

A third reason we may expect a copula approach to benefit ecological
research is that prior work has demonstrated important copula 
stucture in the spatial dependence of environental 
variables [@Serinaldi2008; @She2018; @Goswami2018;
@li2013]. One may expect copula structure in environmental variables 
to take the form of strong tail dependence if
intense meteorological events are also widespread in their effects, as seems 
frequently to be the case.
Spatial dependence in an environmental variable 
tends to beget dependence across space
in fluctuations of populations or other ecosystem variables 
(this is called \emph{spatial synchrony}) which are influenced by the environmental 
variable. This process is called the \emph{Moran effect}. 
If copula structure is transmitted, 
possibly in modified form, 
from environmental to ecological variables, then we may expect important 
copula structure to be a common feature of spatial synchrony of population, community, 
biogeochemical and other ecological variables.
Synchrony is a phenomenon of great interest in 
ecology [@Liebhold2004; @Sheppard2016; @walteretal2017].

@@

We approached our overall goal of exploring the potential of 
copula approaches in ecology by addressing the 
following specific questions. (Q1) Do datasets in ecology and related fields 
have copula structure distinct from that of a Gaussian/normal distribution (here called 
non-normal copula structure)? Do they show tail dependence distinct from that 
of a normal distribution, and in particular do they show asymmetric tail 
dependence? (Q2) If so, what are some possible causes/mechanisms of this 
non-normal copula structure? And (Q3) what are the consequences of non-normal 
copula structure and tail dependence for our understanding of biology and for 
applications? To address Q1, we analyze several datasets, including 
environmental, species-trait, phenological, population, community, and 
ecosystem functioning data. We selected these datasets to span multiple 
sub-fields and organizational levels of ecology. The second section of the 
paper introduces the data, and the third and fourth sections use the data to 
address Q1. The fifth and sixth sections, which address Q2, use simple models to 
explore mechanisms. The seventh and eigth sections, which address Q3, use both
the data of section two as well as models to explore a variety of consequences
of copula structure and tail dependence for applications and for understanding
ecology. The ninth section is the Discussion. 
Copula approaches have been used to great benefit in neuroscience [@onken2009], 
bioinformatics 
[@Kim2008], medical research [@Emura2016], direct study of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018; @li2013], finance, and other fields. 
To our knowledge, these approaches have been used seldom or never in ecology. 
We argue that benefits to ecology of using a complete statistical 
desription of dependence will be substantial.

@@

Datasets included environmental, species-trait, phenological, 
population, community, and ecosystem functioning data. Copula structure describing 
the relationship between atmospheric weather variables such as rainfall or wind speed 
as measured in multiple locations has been 
examined previously in the meteorological literature (e.g., @Serinaldi2008, @li2013). We therefore 
examined environmental variables from the soil instead, using the Rapid Carbon 
Assessment database (RaCA; @Wills2014).
<!--***DAN: Add Wills and Loecke pending reference if published in time--> 
The database comprises measurements of soil organic carbon and total soil 
nitrogen (Mg C or N per hectare of soil surface) from `r nrow(d_soilCN)` 
locations across the coterminous United States (Fig. \ref{SM-fig_SoilCNmapUSA} for sampling locations).
Measurments were from the central pedon of a soil core, to 1 m depth.
<!--***DAN: Terry to check the wording and accuracy here-->
Species-trait data were average species basal metabolic rate (KJ per hour) and 
body mass (grams) data for `r nrow(d_birdsBMR)`  species of birds [@Mcnab_2009] 
and `r nrow(d_mammalsBMR)`  species of mammals [@Mcnab_2008].
These data have been much studied, but to our knowledge the copula structure 
of the dependence between metabolic rate and body mass has not been examined. 
Species-trait data such as these reflect the coevolution of the two traits.
Phenological data were aphid first-flight dates from 10 locations (Fig. \ref{SM-fig_aphid_map}) 
across the United Kingdom (UK) for 20 aphid species (Table \ref{SM-tab_aphid_info}), for the 
35 years 1976 to 2010. These time series were computed from the Rothamsted Insect 
Survey suction-trap dataset [@Harrington2014; @Bell2015]. The first of our two 
population-level datasets was also derived from the RIS suction-trap data, and comprised 
total counts of aphids trapped for the same locations, species, and years; these are 
indicators of aphid migration counts. Data pre-processing for the aphid data was the same as 
that of @Sheppard2016,
but we exlcuded one of the 11 sites they used so as to include only sites with at least 30 
years data.  Our second population dataset comprised average annual abundance estimates 
for 14 locations (Fig. \ref{SM-fig_plankton_map})
in the North Sea and British seas for 22 plankton taxa (Table \ref{SM-tab_plankton_info}) for the 
56 years 1958 to 2013. The locations are $2^\circ$ by $2^\circ$ grid cells. These data 
were obtained from the Continuous Plankton Recorder (CPR) dataset maintained by the Sir 
Alistair Hardy Foundation for Ocean Science (SAHFOS), now part of the Marine Biological 
Association of the UK. Data pre-processing was similar but not identical to that 
of @Sheppard2017; see Appendix \ref{SM-Data_details} for details.
Community-level data were plant aboveground biomass and Shannon's diversity index for `r nrow(d_cc[[5]])` 
plots, each 9m by 9m, maintained at Cedar Creek Ecosystem Science Reserve and sampled in 
the year 2000 [@Tilman2001; @Tilman2006]. 
Finally, ecosystem functioning data were methane (CH$_{4}$) fluxes between the soil or 
water surface and the troposphere, measured at 13 locations at daily to weekly intervals 
from September 2015 to September 2016 at the Great Miami Wetland mitigation bank in Trotwood, 
Ohio [@Holland1999; @Jarecke2016; @smyth2019using]. Each included location was measured on at least 50 dates. 
Sampling locations spanned a hydrologic gradient from upland to seasonal wetland.
See Appendix \ref{SM-Data_details} for additional data details.

@@

<!--summary table with info of all data used in Paper-->
```{r tab_summary_data_info, echo=F, results="asis",message=F}
library(kableExtra)
library(dplyr)
tab_data_info<-as.data.frame(matrix(NA,nrow=8,ncol=4))
colnames(tab_data_info)<-c("Data","Category","No. of measurements we used", "References")
tab_data_info$Data<-c("Soil C and N","Bird masses and BMR", "Mammal masses and BMR", "Cedar Creek data", 
                      "Green spruce aphid abundance", "Leaf-curling plum aphid first flight","Ceratium furca abundance","Methane flux")
tab_data_info$Category<-c("Environmental","Species-trait","Species-trait","Community level",
                          "Population","Phenological","Population","Ecosystem-functioning")
tab_data_info$`No. of measurements we used`<-c(paste(nrow(d_soilCN)," locations",sep=""),
                                               paste(nrow(d_birdsBMR)," birds",sep=""),
                                               paste(nrow(d_mammalsBMR)," mammals",sep=""),
                                               paste(nrow(d_cc[[5]])," plots",sep=""),
                                               "10 locations with atleast 30 yrs. of timeseries for each 20 sp.",
                                               "10 locations with atleast 30 yrs. of timeseries for each 20 sp.",
                                               "14 locations with atleast 45 yrs. of timeseries for each 22 taxa",
                                               "13 locations with atleast 50 dates of data")
tab_data_info$References<-c("US Rapid Carbon Assessment database (RaCA), Wills et al. (2014)", "McNab 2009","McNab 2008",
                            "Cedar Creek Ecosystem Science Reserve",
                            "Rothamsted Insect Survey","Rothamsted Insect Survey","Continuous Plankton Recorder Survey",
                            "Great Miami Wetland Mitigation Bank, Smyth et al. (2019)")

@@

We addressed Q1 principally via a model selection procedure in which the normal copula
the alternative copulas listed above were fit to 
our ecological datasets, and fits of the various copula models 
were compared via the Akaike and Bayesian 
Information Criteria (AIC and BIC). For bivariate datasets 
$(x_i,y_i)$ for $i=1,...,n$, model selection involved several steps. 
First, as in the Introduction, we produced ranks 
$(\tilde{u}_i,\tilde{v}_i)$ where $\tilde{u}_i$ is the rank of $x_i$ in the set 
$\{x_i:i=1,...,n\}$ and $\tilde{v}_i$ is the rank of $y_i$ in the set $\{y_i:i=1,...,n\}$.
Ranks were $1$ for the smallest element in a set. We then 
set $u_i=\tilde{u}_i/(n+1)$ and $v_i=\tilde{v}_i/(n+1)$; subsequent analysis used these 
*normalized ranks* $u_i$ and $v_i$. 
Second, we did a test of the independence of the $u_i$ and $v_i$ 
(implemented in `BiCopIndTest` in the `VineCopula` package). 
Subsequent model 
selection algorithms are ineffective if data could not be distinguished from independent data 
since most or all bivariate copula families we considered include the independent copula in the 
family or as a boundary case. If independence could be rejected ($0.05$ significance level), model 
selection proceeded. Third, we fit the bivariate copula families listed previously to the ranked data 
$(u_i,v_i)$ via maximum likelihood, using the `BiCopEst` function in `VineCopula`, and 
obtained AIC and BIC values and accompanying model weights, $\text{AIC}_{\text{w}}$ and 
$\text{BIC}_{\text{w}}$ [@burnham2003_modelselection]. The weight of a model represents the 
relative support of data for the model; weights sum to 1 across models considered. The lower- and 
upper-tail dependence of the best-fitting member of each family was also provided by `BiCopEst`. 
Fourth, $\text{AIC}_{\text{w}}$ values for all fitted models were used to get
model-averaged lower- and upper-tail dependence values using standard model averaging 
formulas [@burnham2003_modelselection]; likewise for BIC.
Model selection methods will give the *relative* support of several models, but will not,
on their own, indicate whether any of the models selected is an objectively good fit: all 
models considered might poorly describe the data. To test this, and as the final step of
our procedure, we tested the goodness of fit of our AIC-best copula family using a 
bootstrapping procedure implemented in `BiCopGofTest` of `VineCopula`. The method performed 
two tests, one based on a Cramer-von Mises statistic, and one based on a 
Kolmogorov-Smirnov statistic. To keep computation times reasonable, 
an initial run using $100$ bootstraps was performed, and if
the $p$-value resulting from either test was less than $0.2$, tests were re-run with 
$1000$ boostraps.

@@

\begin{figure}
\begin{center}
\includegraphics[width=6.5in]{./Results/BivarDataPlot.pdf}
\caption{Upper panels show raw data plots for (A) soil C and N data, (B) $\log_{10}$(basal metabolic rate (BMR)) vs. $\log_{10}$(body mass) 
for birds, (C) $\log_{10}$(BMR) vs. $\log_{10}$(body mass) for mammals, and (D) above-ground plant biomass vs. Shannon's index (H) from Cedar Creek. Bottom panels show the corresponding copula plots. See text for units used in upper panels. \label{fig_biv_multi_raw_cop}}
\end{center}
\end{figure}

@@

<!--multivariate summary-->
Table \ref{tab_multivar_summary} summarizes some of our results using our
multivariate data. In addition to summaries of the green spruce aphid abundance data results 
elaborated in detail above, the table also summarizes results for 
leaf-curling plum aphid first flight data, abundance data for the dinoflagellate
\textit{Ceratium furca}, and methane-flux data (see Data). Results supported the 
conclusions that non-N copula structure, non-N tail dependence, and asymmetric tail 
dependence were common. Most locations pairs were non-independent 
(Table \ref{tab_multivar_summary}, row 2). The large majority of non-independent location 
pairs had best-fitting copula that was not the N copula (Table \ref{tab_multivar_summary}, 
row 3), and AIC weights of best-fitting copulas were 
substantially better than AIC weights of N copulas (Table \ref{tab_multivar_summary}, rows 4-5).
For these data, best-fitting copulas were nearly always considered an adequate fit, 
according to Cramer-von Mises and Kolmogorov-Smirnov goodness of fit tests 
(Table \ref{tab_multivar_summary}, row 6). Some of the datasets (green spruce aphid
abundance, \textit{Ceratium furca} abundance, methane flux) showed stronger 
or marginally stronger lower- than upper-tail dependence 
(Table \ref{tab_multivar_summary}, rows 7-9), whereas leaf-curling plum aphid first flight
data showed the reverse. Asymmetry results were verified by nonparametric approaches 
<!--(e.g., Table \ref{tab_multivar_summary}, row 10)-->
(Appendix \ref{SM-Q1multvarnonparam}). Detailed results are in 
Tables \ref{SM-tab_count_bestcop}-\ref{SM-tab_methane_resamp} and 
Figs \ref{SM-fig_aphid_count}-\ref{SM-fig_methane}.
We also carried out the same analyses for abundance and first-flight data for the 18 
aphid species for which we had data other than the green spruce and leaf-curling plum aphids,
as well as for the 21 plankton taxa for which we had abundance data other than the
\textit{Ceratium furca} (results not shown). The conclusions were generally supported that 
non-N copula structure and tail dependence, as well as asymmetry of tail dependence, are 
very common.

@@

knitr::kable(tab_multivar_summary[1:9,], 
             format="latex", escape = F,
             caption.short = "Summary of results for multivariate datasets",
             caption = "Summary of results for multivariate datasets. Rows 3-9 of the table were computed for the non-independent pairs (row 2) only. A well-fitting location pair (row 6) was one for which the best-fitting copula had $p$-values $>0.01$ for both the Cramer-von Mises and Kolmogorov-Smirnov goodness of fit tests. \\label{tab_multivar_summary}",
             booktabs=T, linesep = "\\addlinespace",
             col.names = c(" ","Green spruce aphid abundance",
                        "Leaf-curling plum aphid first flight",
                        "Ceratium furca abundance","Methane flux")
             )%>%column_spec(2:5,width="2.5cm")%>%
             column_spec(1, width = "7cm")%>%kable_styling(latex_options = "basic",position = "center")
```

>>>>

---
title: "The importance of a complete statistical description of dependence between variables for ecology and related fields"
author: "Shyamolina Ghosh, Lawrence W. Sheppard, Mark T. Holder, Terrance E. Loecke, Philip C. Reid, Daniel C. Reuman"
date: ""
geometry: "left=1.5cm,right=1.5cm,top=2.5cm,bottom=2.8cm"

@@

Reid: School of Biological & Marine Sciences, University of Plymouth, Drake Circus, Plymouth PL4 8AA, United Kingdom

@@

Very generally, all branches of ecology and related areas <!--DAN : Reid said " Is this text needed. 
Ecology covers 
all branches of the biological sciences?
What do you mean by related areas? 
You use related fields later.
Needs clarifying."-->
study relationships among environmental and biological 
variables. However, commonly used correlation and regression 
approaches to studying 
such relationships are limited - such statistical
approaches provide only a very partial description of a relationship between variables. 
For instance, datasets showing markedly different 
relationships may nevertheless have the same correlation coefficient value 
(Fig. \ref{fig_cop_pedag1}A-B). The variables of Fig. \ref{fig_cop_pedag1}A 
(respectively, Fig. \ref{fig_cop_pedag1}B) are principally 
related in the left (respectively, right) portions of their distributions, an asymmetric 
pattern of dependence that may
have ecological significance, as discussed below, but is not captured by 
correlation. Panels C and D 
were obtained from A and B through monotonic transformation 
(Appendix \ref{SM-pedagog1_details} for details),
which alters the Pearson, but not the Spearman correlation. Although correlation
is only one of numerous metrics for 
studying relationships between variables, other common metrics <!--DAN : add example--> also 
provide only partial information.

@@

Other well-developed statistical tools exist, that are 
applied widely in other fields (e.g. finance, neuroscience, bioinformatics etc.), 
and provide a demonstrably complete 
description of the dependence 
between two or more variables [@joe2014_dependence; @nelsen2006_copula]. 
The goal of this paper is to illustrate the potential for 
broader application of these
tools in ecology and related fields. The tools are based on 
the statistical concept of the *copula*, which we here 
introduce. Copula approaches separate the information contained in a bivariate 
random variable into two distinct parts: the information contained in the 
marginal distributions (which says nothing about the dependence between the 
variables), and the remaining information (which is solely about
the dependence between the variables). From a data perspective, given 
a sample $(x_i,y_i)$, $i=1,2,\ldots,n$ (e.g., 
Fig. \ref{fig_cop_pedag1}A-B), the associated dependence
or copula structure is revealed by the bivariate plot of $u_i$ versus $v_i$, where
$u_i$ is the rank of $x_i$ in the set of the $x_j$ ($j=1,\ldots,n$), divided by
$n+1$; and $v_i$ is defined in the same way but using the $y_i$. See, 
for instance,
Fig. \ref{fig_cop_pedag1}E-F, which show the copula structure of 
A-B, respectively. Ranking serves to uniformize the 
marginal distributions of the underlying random variables, isolating the 
dependence structure only.
Mathematically, if $(X,Y)$ is a continuous bivariate random
variable and the cumulative distribution functions (cdfs) of $X$ and $Y$ are 
$F_X(x)=P[X\leq x]$ and $F_Y(y)=P[Y\leq y]$, then the dependence or copula structure
of $(X,Y)$ is revealed by the random variable $(F_X(X),F_Y(Y))$.
This random variable takes values in the unit square, 
$[0,1] \times [0,1]$, and has uniform marginal distributions. Probability density 
functions (pdfs) corresponding to the copulas associated 
with Fig. \ref{fig_cop_pedag1}E-F are in Fig.
\ref{fig_cop_pedag1}G-H, respectively. 
See @nelsen2006_copula and @joe2014_dependence
for general background. A theorem of A. @sklar1959_theorem
states that copulas give a unique and complete 
representation of the dependence
between variables, in the sense that the joint distribution of $(X,Y)$ 
can be reconstructed from the 
copula and the marginal distributions. 
Copula theory is also well developed for
multivariate distributions [@joe1997multivariate].

@@

A third reason we may expect a copula approach to benefit ecological
research is that prior work has demonstrated important copula 
stucture in the spatial dependence of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018;
@li2013]. One may expect copula structure in environmental variables 
to take the form of strong tail dependence if
intense meteorological events are also widespread in their effects, as seems 
frequently to be the case.
Spatial dependence in an environmental variable 
tends to beget dependence across space
in fluctuations of populations or other ecosystem variables 
(this is called \emph{spatial synchrony}) which are influenced by the environmental 
variable. This process is called the \emph{Moran effect}. 
If copula structure is transmitted, 
possibly in modified form, 
from environmental to ecological variables, then we may expect important 
copula structure to be a common feature of spatial synchrony of population, community, 
biogeochemical and other ecological variables.
Synchrony is a phenomenon of great interest in 
ecology [@Liebhold2004; @Sheppard2016; @walteretal2017].

@@

We approached our overall goal of exploring the potential of 
copula approaches in ecology by addressing the 
following specific questions. (Q1) Do datasets in ecology and related fields 
have copula structure distinct from that of a Gaussian/normal distribution (here called 
non-normal copula structure)? Do they show tail dependence distinct from that 
of a normal distribution, and in particular do they show asymmetric tail 
dependence? (Q2) If so, what are some possible causes/mechanisms of this 
non-normal copula structure? And (Q3) what are the consequences of non-normal 
copula structure and tail dependence for our understanding of biology and for 
applications? To address Q1, we analyze several datasets, including 
environmental, species-trait, phenological, population, community, and 
ecosystem functioning data. We selected these datasets to span multiple 
sub-fields and organizational levels of ecology. The second section of the 
paper introduces the data, and the third and fourth sections use the data to 
address Q1. The fifth and sixth sections, which address Q2, use simple models to 
explore mechanisms. The seventh and eigth sections, which address Q3, use both
the data of section two as well as models to explore a variety of consequences
of copula structure and tail dependence for applications and for understanding
ecology. The ninth section is the Discussion. 
Copula approaches have been used to great benefit in neuroscience [@onken2009], 
bioinformatics 
[@Kim2008], medical research [@Emura2016], direct study of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018; @li2013], finance, and other fields. 
To our knowledge, these approaches have never or seldom been used for ecological research. 
We argue that benefits to ecology of using a complete statistical 
desription of dependence will be substantial.

@@

Datasets included environmental, species-trait, phenological, 
population, community, and ecosystem functioning data. Copula structure describing 
the relationship between atmospheric weather variables such as rainfall or wind speed 
as measured in multiple locations has been 
examined previously in the meteorological literature [e.g., @Serinaldi2008; @li2013]. We therefore 
examined environmental variables from the soil instead, using the Rapid Carbon 
Assessment database [RaCA; @Wills2014].
<!--***DAN: Add Wills and Loecke pending reference if published in time--> 
The database comprises measurements of soil organic carbon and total soil 
nitrogen (Mg C or N per hectare of soil surface) from `r nrow(d_soilCN)` 
locations across the coterminous United States (Fig. \ref{SM-fig_SoilCNmapUSA} for sampling locations).
Measurments were from the central pedon of a soil core, to 1 m depth.
<!--***DAN: Terry to check the wording and accuracy here-->
Species-trait data were average species basal metabolic rate (KJ per hour) and 
body mass (grams) data for `r nrow(d_birdsBMR)`  species of birds [@Mcnab_2009] 
and `r nrow(d_mammalsBMR)`  species of mammals [@Mcnab_2008].
These data have been much studied, but to our knowledge the copula structure 
of the dependence between metabolic rate and body mass has not been examined. 
Species-trait data such as these reflect the coevolution of the two traits.
Phenological data were aphid first-flight dates from 10 locations (Fig. \ref{SM-fig_aphid_map}) 
across the United Kingdom (UK) for 20 aphid species (Table \ref{SM-tab_aphid_info}), for the 
35 years 1976 to 2010. These time series were computed from the Rothamsted Insect 
Survey (RIS) suction-trap dataset [@Harrington2014; @Bell2015]. The first of our two 
population-level datasets was also derived from the RIS suction-trap data, and comprised 
total counts of aphids trapped for the same locations, species, and years; these are 
indicators of aphid migration counts. Data pre-processing for the aphid data was the same as 
that of @Sheppard2016,
but we excluded one of the 11 sites they used in order to only use sites with at least 30 
years data.  Our second population dataset comprised average annual abundance estimates 
for 14 locations (Fig. \ref{SM-fig_plankton_map})
in the North Sea and British seas for 22 plankton taxa (Table \ref{SM-tab_plankton_info}) for the 
56 years 1958 to 2013. The locations are $2^\circ$ by $2^\circ$ grid cells. These data 
were obtained from the Continuous Plankton Recorder (CPR) dataset now operated and maintained by 
the Marine Biological Association. Data pre-processing was similar but not identical to that 
of @Sheppard2017; see Appendix \ref{SM-Data_details} for details.
Community-level data obtained from the Cedar Creek Ecosystem Science Reserve were <!--DAN : check this rephrasing-->
plant aboveground biomass and plant percent cover data (we used this percent cover data to calculate 
Shannon's diversity index) for `r nrow(d_cc[[5]])` plots, each 9m by 9m, as sampled in 
the year 2000 [@Tilman2001; @Tilman2006] <!--actually they collected data in each year of 1996,1997,1998,1999,2000,2007-->. 
Finally, ecosystem functioning data were methane (CH$_{4}$) fluxes between the soil or 
water surface and the troposphere, measured at 13 locations at daily to weekly intervals 
from September 2015 to September 2016 at the Great Miami Wetland mitigation bank in Trotwood, 
Ohio [@Holland1999; @Jarecke2016; @smyth2019using]. Each included location was measured on at least 50 dates. 
Sampling locations spanned a hydrologic gradient from upland to seasonal wetland.
See Appendix \ref{SM-Data_details} for additional data details.

@@

<!--summary table with info of all data used in Paper-->
```{r tab_summary_data_info, echo=F, results="asis",message=F}
library(kableExtra)
library(dplyr)
tab_data_info<-as.data.frame(matrix(NA,nrow=8,ncol=4))
colnames(tab_data_info)<-c("Data","Category","No. of measurements we used", "References")
tab_data_info$Data<-c("Soil C and N","Bird masses and BMR", "Mammal masses and BMR", "Cedar Creek data", 
                      "Green spruce aphid abundance", "Leaf-curling plum aphid first flight","Ceratium furca abundance","Methane flux")
tab_data_info$Category<-c("Environmental","Species-trait","Species-trait","Community level",
                          "Population","Phenological","Population","Ecosystem-functioning")
tab_data_info$`No. of measurements we used`<-c(paste(nrow(d_soilCN)," locations",sep=""),
                                               paste(nrow(d_birdsBMR)," birds",sep=""),
                                               paste(nrow(d_mammalsBMR)," mammals",sep=""),
                                               paste(nrow(d_cc[[5]])," plots",sep=""),
                                               "10 locations with atleast 30 yrs. of timeseries for each 20 sp.",
                                               "10 locations with atleast 30 yrs. of timeseries for each 20 sp.",
                                               "14 locations with atleast 45 yrs. of timeseries for each 22 taxa",
                                               "13 locations with atleast 50 dates of data")
tab_data_info$References<-c("US Rapid Carbon Assessment database (RaCA), Wills et al. (2014)", "McNab 2009","McNab 2008",
                            "Cedar Creek Ecosystem Science Reserve [biodiversity experiment e120]",
                            "Rothamsted Insect Survey","Rothamsted Insect Survey","Continuous Plankton Recorder Survey",
                            "Great Miami Wetland Mitigation Bank, Smyth et al. (2019)")

@@

We addressed Q1 principally via a model selection procedure in which the normal copula
the alternative copulas listed above were fitted to 
our ecological datasets. The resulting fits of the various copula models 
were compared via the Akaike and Bayesian 
Information Criteria (AIC and BIC). For bivariate datasets 
$(x_i,y_i)$ for $i=1,...,n$, model selection involved several steps. 
First, as in the Introduction, we produced ranks 
$(\tilde{u}_i,\tilde{v}_i)$ where $\tilde{u}_i$ is the rank of $x_i$ in the set 
$\{x_i:i=1,...,n\}$ and $\tilde{v}_i$ is the rank of $y_i$ in the set $\{y_i:i=1,...,n\}$.
Ranks were $1$ for the smallest element in a set. We then 
set $u_i=\tilde{u}_i/(n+1)$ and $v_i=\tilde{v}_i/(n+1)$; subsequent analysis used these 
*normalized ranks* $u_i$ and $v_i$. 
Second, we did a test of the independence of the $u_i$ and $v_i$ 
(implemented in `BiCopIndTest` in the `VineCopula` package). 
Subsequent model 
selection algorithms are ineffective if data could not be distinguished from independent data 
since most or all bivariate copula families we considered include the independent copula in the 
family or as a boundary case. If independence could be rejected ($0.05$ significance level), model 
selection proceeded. Third, we fit the bivariate copula families listed previously to the ranked data 
$(u_i,v_i)$ via maximum likelihood, using the `BiCopEst` function in `VineCopula`, and 
obtained AIC and BIC values and accompanying model weights, $\text{AIC}_{\text{w}}$ and 
$\text{BIC}_{\text{w}}$ [@burnham2003_modelselection]. The weight of a model represents the 
relative support of data for the model; weights sum to 1 across models considered. The lower- and 
upper-tail dependence of the best-fitting member of each family was also provided by `BiCopEst`. 
Fourth, $\text{AIC}_{\text{w}}$ values for all fitted models were used to get
model-averaged lower- and upper-tail dependence values using standard model averaging 
formulas [@burnham2003_modelselection]; likewise for BIC.
Model selection methods will give the *relative* support of several models, but will not,
on their own, indicate whether any of the models selected is an objectively good fit: all 
models considered might poorly describe the data. To test this, and as the final step of
our procedure, we tested the goodness of fit of our AIC-best copula family using a 
bootstrapping procedure implemented in `BiCopGofTest` of `VineCopula`. The method performed 
two tests, one based on a Cramer-von Mises statistic, and one based on a 
Kolmogorov-Smirnov statistic. To keep computation times reasonable, 
an initial run using $100$ bootstraps was performed, and if
the $p$-value resulting from either test was less than $0.2$, tests were re-run with 
$1000$ boostraps.

@@

\begin{figure}
\begin{center}
\includegraphics[width=6.5in]{./Results/BivarDataPlot.pdf}
\caption{Upper panels show raw data plots for (A) soil C and N data, (B) $\log_{10}$(basal metabolic rate (BMR)) vs. $\log_{10}$(body mass) 
for birds, (C) $\log_{10}$(BMR) vs. $\log_{10}$(body mass) for mammals, and (D) above-ground plant biomass vs. Shannon's index \textit{(H)} from Cedar Creek. Bottom panels show the corresponding copula plots. See text for units used in upper panels. \label{fig_biv_multi_raw_cop}}
\end{center}
\end{figure}

@@

<!--multivariate summary-->
Table \ref{tab_multivar_summary} summarizes some of our results using our
multivariate data. In addition to summaries of the green spruce aphid abundance data results 
elaborated in detail above, the table also summarizes results for 
leaf-curling plum aphid first flight data, abundance data for the dinoflagellate
\textit{Ceratium furca}, and methane-flux data (see Data). Results supported the 
conclusions that non-N copula structure, non-N tail dependence, and asymmetric tail 
dependence were common. Most locations pairs were non-independent 
(Table \ref{tab_multivar_summary}, row 2). The large majority of non-independent location 
pairs had best-fitting copula that was not the N copula (Table \ref{tab_multivar_summary}, 
row 3), and AIC weights of best-fitting copulas were 
substantially better than AIC weights of N copulas (Table \ref{tab_multivar_summary}, rows 4-5).
For these data, best-fitting copulas were nearly always considered an adequate fit, 
according to Cramer-von Mises and Kolmogorov-Smirnov goodness of fit tests 
(Table \ref{tab_multivar_summary}, row 6). Some of the datasets (green spruce aphid
abundance, \textit{Ceratium furca} abundance, methane flux) showed stronger 
or marginally stronger lower- than upper-tail dependence 
(Table \ref{tab_multivar_summary}, rows 7-9), whereas leaf-curling plum aphid first flight
data showed the reverse. Asymmetry results were verified by nonparametric approaches 
<!--(e.g., Table \ref{tab_multivar_summary}, row 10)-->
(Appendix \ref{SM-Q1multvarnonparam}). Detailed results are in 
Tables \ref{SM-tab_count_bestcop}-\ref{SM-tab_methane_resamp} and 
Figs \ref{SM-fig_aphid_count}-\ref{SM-fig_methane}.
We also carried out the same analyses for abundance and first-flight data for the 18 
aphid species for which we had data other than the green spruce and leaf-curling plum aphids,
as well as for the 21 plankton taxa for which we had abundance data other than the
\textit{Ceratium furca} (results not shown). The conclusions were generally supportive that 
non-N copula structure and tail dependence, as well as asymmetry of tail dependence, are 
very common.

@@

knitr::kable(tab_multivar_summary[1:9,], 
             format="latex", escape = F,
             caption.short = "Summary of results for multivariate datasets",
             caption = "Summary of results for multivariate datasets. Rows 3-9 of the table were computed for the non-independent pairs (row 2) only. A well-fitting location pair (row 6) was one for which the best-fitting copula had $p$-values $>0.01$ for both the Cramer-von Mises and Kolmogorov-Smirnov goodness of fit tests. \\label{tab_multivar_summary}",
             booktabs=T, linesep = "\\addlinespace",
             col.names = c(" ","Green spruce aphid abundance",
                        "Leaf-curling plum aphid first flight",
                        "\\textit{Ceratium furca} abundance","Methane flux")
             )%>%column_spec(2:5,width="2.5cm")%>%
             column_spec(1, width = "7cm")%>%kable_styling(latex_options = "basic",position = "center")
```
